Rigidity of Expectations: Additional Evidence from Density Forecasts of Professionals and Households

### 2020-04-02 paper/InfVar.tex
drafted the codes of GMM for SE in GMM_Dev. Needed to test.
started revission drafting. finished first draft of the introduction.

%opening
\title{Rigidity of Expectations: Additional Evidence from Density Forecasts of the Inflation by Professionals and Households}
\author{Tao Wang \footnote{Department of Economics, Johns Hopkins University; email: twang80@jhu.edu. }}

@@

	Theories on how agents form expectations in ways deviating from rational expectation(RE) have proliferated over the past decade. This has been also accompanied by a welcoming recognition of the usefulness of surveys by macroeconomists in testing these theories empirically.  On one hand, various theories built upon different micro-foundations produce somewhat similar macro patterns. For instance, the notion of expectation rigidity micro-founded by different theories mostly generates a sluggish response to new information.  On the other hand, there are important and subtle differences in testable predictions from these theories for both individual forecasts and aggregate moments of the forecast.

	This paper extends this line of effort by bringing an additional moment of forecasts into perspective, i.e. the uncertainty. Existing work that studies expectation formation utilizing survey data mostly focus on the individual and cross-sectional patterns of the mean forecast of the variable. But there are additional insights from the forecast uncertainty, which is only available if each surveyee is asked to assign their own perceived probabilities to a range of values of the variable\footnote{For a survey on the importance of measuring subjective expectations using probabilistic surveys, see \citet{manski2004measuring}. See \citet{manski2018survey} for the recent development of the discussion.}. For instance, the dispersion of uncertainty across individual forecasters, as that of the average forecasts, are inconsistent with the benchmark full-information rational expectation as the latter assumes that agents agree on the generating process of the data and have the common knowledge of available information. Dynamically, across different vintages of the forecast, the revision in uncertainty is a measure of information gain or the degree of forecasting efficiency. Besides, the relationship between uncertainty with other moments such as forecast errors and disagreements provides a way to check if the surveyed expectations are consistent with certain theories of expectation formation. Using these insights, the paper finds additional evidence from uncertainty consistent with the rigidity models. In the same time, the degree of rigidity appears to be shock-specific and bears the difference between households and professional forecasters.

	There are three major contributions of this paper. On the theoretical side, I explicitly derive the prediction of various theories of expectation formation about the dynamics of uncertainty. For instance,  different theories that entail rigidities in incorporating new information predict the inefficiency in forecast revisions. This paper shows that this inefficiency can not only be seen in the average forecast, but also in the forecast uncertainty. This imposes additional restrictions on parametric configurations that produce the degree of rigidity seen in the data. Specifically, I find that the configurations of parameters in sticky expectation models are more realistic than that in noisy information theories.

	On the empirical side, the contribution is that we provide additional tests of the null of full-information rational expectation using the revision of uncertainty. I first replicate the results of tests based on forecast errors using microdata of households and professionals following the existing literature. Then, I undertake autoregression-based tests for revisions in mean forecasts and uncertainty, respectively. The new result from this paper is that the revision in uncertainty has a serial correlation that is not consistent with the level of forecast efficiency predicted by rational expectation. This provides an additional result that rejects the null of the rational expectation hypothesis.

	In addition, using externally identified shocks to inflation, i.e. productivity shocks, oil price shocks, and monetary policy shocks, I find evidence that the degree of rigidity is shock-specific. In particular, professionals seem to be responsive to monetary policy shocks, while households are more responsive to oil price shocks. Unlike testing the null-hypothesis, the externally identified shocks allow us to look into the dynamic impulse responses of different moments of expectations to the particular shock. This can be directly compared with the predictions of the theories. I follow the framework by \citet{coibion2012can} while making two variations. First, I do not only examine the dynamics of forecast error, disagreement, but also the forecast uncertainty. Second, I include two monetary policy shocks, the shocks to current federal fund rates and future path of the federal funds rate.

	This paper is related to four strands of literature. First, it is realted to a series of empirical work directly testing and evaluating  various theories on expectations formation using survey data. For instance, \citet{mankiw2003disagreement}, \citet{carroll2003macroeconomic}, \citet{branch2004theory}. More recent examples include \citet{coibion2018firms} on firms' managers. In addition to testing particular sets of theories, there is also a number of papers that show people's expectations are driven by idiosyncratic demographics, cognitive abilities and macroeconomic histories experienced(\citet{malmendier2015learning}, \citet{das2017socioeconomic} and \citet{d2019iq}, etc.). In terms of the methodology, this paper is closest to \citet{coibion2012can} and \citet{fuhrer2018intrinsic}.  However, all of these studies simply rely upon point forecasts instead of density forecast or surveyed uncertainty. This is one theme on which this paper differs from the existing literature.

@@

@@

	\begin{eqnarray}\label{NoPastInfFE}
		\overline{FE}^{*}_{i,t+h|t} = - \sum^{h}_{k=1} \rho^k \omega_{t+k}
	\end{eqnarray}

@@

	\begin{eqnarray}
		\begin{aligned}
			\Sigma^y_{i,t|t} = \Sigma^y_{i,t|t-1} - \Sigma^y_{i,t|t-1} H'(H \Sigma^y_{i,t-1} H' +\Sigma^v) H \Sigma^y_{i,t|t-1} 
		\end{aligned}
	\end{eqnarray}

@@

	\begin{eqnarray}\label{VarNIRv}
		\Sigma^y_{i,t|t} - \Sigma^y_{i,t|t-1} = - \Sigma^y_{i,t|t-1} H'(H \Sigma^y_{i,t-1} H' +\Sigma^v) H \Sigma^y_{i,t|t-1} <0
	\end{eqnarray}

>>>>

%opening
\title{How do agents form inflation expectations? evidences from cross-moment estimation and the uncertainty}
\author{Tao Wang \footnote{Department of Economics, Johns Hopkins University; email: twang80@jhu.edu. }}

@@

	Theories on how agents form expectations in ways deviating from rational expectations (RE) have proliferated over the past decade. This has been facilitated by a welcoming recognition of the usefulness of surveys by macroeconomists in testing these theories empirically. On one hand, various theories built upon different micro-foundations produce somewhat similar macro patterns. For instance, the notion of expectation rigidity micro-founded by different theories mostly generates a sluggish response to new information. On the other hand, there are important and subtle differences in testable predictions from these theories for both individual forecasts and aggregate moments of the forecast.  The seminal work by \cite{coibion2012can} examines predictions from different theories in a comparable manner using historical data on inflation expectations. The authors derive testable predictions for a variety of theory and exmains the impulse responses of the inflation expectations to different externally estimated shocks to inflation.

	This paper extends the abovementioned effort by estimating each theory parametrically accounting for cross-moment restrictions. This is a natural step forward from reduced-form tests of rationality widely seen in the literature. With comparable estimates of different theories, one can evaluate to what extent each theory fits the observed expectation data and inflation dynamics.  I evaluate each model of expectation in terms of its sensitivity of specification with respect to four criteria (1) what moments are used for estimation. (2) the specification of the underlying process of the inflation. (3)  if estimating the underlying process and expectations separately or jointly. The former basicaly assumes recovers the inflation process only based on inflation data while the later let the expectations to provide information used for inflation estimation as well. (4) if it accounts for households and forecasters' expectations equally well.  Broadly speaking, this framework is closer to the original spirit of rational expectation in the sense it focuses on the consistency of the model specification instead of the absolute rationality.

	This is not the first research that estimates theory using cross-moment restrictions (For instance, \cite{giacomini2020heterogeneity}). But one novel aspect of this paper lies in utilizing an additional moment of forecasts in estimation, i.e. the uncertainty. Existing work that studies expectation formation utilizing survey data mostly focuses on the individual and cross-sectional patterns of the mean forecast of the variable. But there are additional insights from the forecast uncertainty, which is only available if each surveyee is asked to assign their own perceived probabilities to a range of values of the variable. For instance, the dispersion of uncertainty across individual forecasters, as that of the average forecasts, are inconsistent with the benchmark full-information rational expectation as the latter assumes that agents agree on the generating process of the data and have the common knowledge of available information. Dynamically, across different vintages of the forecast, the revision in uncertainty is a measure of information gain or the degree of forecasting efficiency. Besides, the relationship between uncertainty with other moments such as forecast errors and disagreements provides a way to check if the surveyed expectations are consistent with certain theories of expectation formation. In addition, although some theories have similar qualitative predictions about the forecasting moments, the configurations of the model parameters consistent with the theory may not be empirically realistic.

	With the framework of estimation presented above, this paper evaluates three workhorse theories on expectation formation. (1) Sticky expectation (SE). (2) Noisy information (NI). (3) Diagnostic expectations (DE).  One major distinction between the first two and DE is that the former predict underreaction due to rigidity and the latter predicts overreaction.  The distinction between SE and NI subtly depends on the quantitative realism of the data implied parameters.

	The findings from the paper are the following.
	Overall, sticky expectation (SE) augmented with inflation of stochastic volatility (SV) fits the joint dynamics of inflation expectation and inflations better than other theories, for both households and professionals. 
	Three theories perform indistinguishably well in matching dynamics of forecast errors but differ in matching dynamics of disagreement and uncertainty. 
	All of the three theories' parameter estimates are not particularly sensitive to the use of moments in estimation. 
	Both SE and DE perform well in that the parameter estimate is not sensitive to if allowing for expectation data to be used to estimate the inflation process itself. In contrast, the NI coefficient estimate changes substantially if the process of inflation is jointly done, particularly for households. 
	In terms of sensitivity to the specification of the inflation process, SE and DE have done a relatively good job for professionals but less consistent for households. NI estimates are very sensitive to the assumption about the inflation process.
	Compared to professional forecasters, households show a limited degree of consistency within many theories. This evidence of inconsistency adds to the evidence rejecting full-information-rationality based on reduced-form tests.

	Because this paper adds uncertainty to the moments used for estimation, another contribution of this paper is that I derive explicitly prediction of various theories of expectation formation about the dynamics of uncertainty. For instance, different theories that entail rigidities in incorporating new information predict the inefficiency in forecast revisions. This paper shows that this inefficiency can not only be seen in the average forecast, but also in the forecast uncertainty. This imposes additional restrictions on parametric configurations that produce the degree of rigidity seen in the data.

	On the empirical side, besides structural estimation, I also provide additional tests of the null of full-information rational expectation using the revision of uncertainty. I first replicate the results of tests based on forecast errors using microdata of households and professionals following the existing literature. Then, I undertake autoregression- based tests for revisions in mean forecasts and uncertainty, respectively. The new result from this paper is that the revision in uncertainty has a serial correlation that is not consistent with the level of forecast efficiency predicted by rational expectation. This provides an additional result that rejects the null of the rational expectation hypothesis.

	In addition, using externally identified shocks to inflation, i.e. productivity shocks, oil price shocks, and monetary policy shocks, I find evidence that the degree of rigidity is shock-specific. In particular, professionals seem to be responsive to monetary policy shocks, while households are more responsive to oil price shocks. Unlike testing the null-hypothesis, the externally identified shocks allow us to look into the dynamic impulse responses of different moments of expectations to the particular shock. This can be directly compared with the predictions of the theories. I follow the framework by \cite{coibion2012can} while making two variations. First, I do not only examine the dynamics of forecast error, disagreement, but also the forecast uncertainty. Second, I include two monetary policy shocks, the shocks to current federal fund rates and future path of the federal funds rate. 
	Related literature

	\subsection{Related Literature}
	This paper is related to four strands of literature. First, it is realted to a series of empirical work directly testing and evaluating  various theories on expectations formation using survey data. For instance, \citet{mankiw2003disagreement}, \citet{carroll2003macroeconomic}, \citet{branch2004theory}. More recent examples include \citet{coibion2018firms} on firms' managers. In addition to testing particular sets of theories, there is also a number of papers that show people's expectations are driven by idiosyncratic demographics, cognitive abilities and macroeconomic histories experienced(\citet{malmendier2015learning}, \citet{das2017socioeconomic} and \citet{d2019iq}, etc.). In terms of the methodology, this paper is closest to \cite{giacomini2020heterogeneity}, which estimates theories of expectation formation using cross-moments restrictions. However, all of these studies simply rely upon point forecasts instead of density forecast or surveyed uncertainty. This is one theme on which this paper differs from the existing literature.

@@

	%%%%%%%%%%%%%%%%%

@@

	\begin{eqnarray}\label{NoPastInfFE}
		\overline{FE}^{*}_{i,t+h|t} = - \sum^{h-1}_{k=0} \rho^k \omega_{t+h-k}
	\end{eqnarray}

@@

	\begin{eqnarray}
		\begin{aligned}
			\Sigma^y_{i,t|t} = \Sigma^y_{i,t|t-1} - \Sigma^y_{i,t|t-1} H'{(H \Sigma^y_{i,t-1} H' +\Sigma^v)}^{-1} H \Sigma^y_{i,t|t-1} 
		\end{aligned}
	\end{eqnarray}

@@

	\begin{eqnarray}\label{VarNIRv}
		\Sigma^y_{i,t|t} - \Sigma^y_{i,t|t-1} = - \Sigma^y_{i,t|t-1} H'{(H \Sigma^y_{i,t-1} H' +\Sigma^v)}^{-1} H \Sigma^y_{i,t|t-1} <0
	\end{eqnarray}

### 2020-04-02 slides/ExpSlides.tex
started revission drafting. finished first draft of the introduction.

\documentclass{beamer}
\usepackage{default}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{adjustbox}  % Allows for fitting tables into slide
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage{caption}
%\usepackage{subcaption}
\usepackage{natbib}
\usepackage{adjustbox}
\usepackage{subcaption}
%\usetheme{AnnArbor}
%\usetheme{Antibes}
%\usetheme{Bergen}
%\usetheme{Berkeley}
%\usetheme{Berlin}
\usetheme{Boadilla}
%\usetheme{boxes}
%\usetheme{CambridgeUS}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{default}
%\usetheme{Frankfurt}
%\usetheme{Goettingen}
%\usetheme{Hannover}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Luebeck}
%\usetheme{Madrid}
%\usetheme{Malmoe}
%\usetheme{Marburg}
%\usetheme{Montpellier}
%\usetheme{PaloAlto}
%\usetheme{Pittsburgh}
%\usetheme{Rochester}
%\usetheme{Singapore}
%\usetheme{Szeged}
%\usetheme{Warsaw}

@@

\begin{frame}{Scoring card for a theory of expectation formation}
	To look if the parameter and fitness is relatively robust to 
	\begin{enumerate}
		\item use of different moments in estimation
		\item alternative assumption about underlying process 
		\item two-step estimation or joint-estimation 
		\item relatively fit with professionals and households  
	\end{enumerate}
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%
\subsubsection{Hybird of NI and SE}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{frame}{Conclusion}
	\begin{itemize}
		\item Sticky expectation (SE) matches data of inflation and expectations better compared to noisy information (NI) 
		\item Within each model,  households are more irrational compared to professionals
		\item Incorporating higher moments, i.e. uncertainty,  helps ``discipline'' theories on expectation formation
		\item  Higher moments from surveys also contain useful information about the inflation dynamics itself
	\end{itemize}	
\end{frame}

>>>>

\documentclass{beamer}
\usepackage{default}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{adjustbox}  % Allows for fitting tables into slide
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage{caption}
%\usepackage{subcaption}
\usepackage{natbib}
\usepackage{adjustbox}
\usepackage{subcaption}
\usepackage{verbatim}

%\usetheme{AnnArbor}
%\usetheme{Antibes}
%\usetheme{Bergen}
%\usetheme{Berkeley}
%\usetheme{Berlin}
\usetheme{Boadilla}
%\usetheme{boxes}
%\usetheme{CambridgeUS}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{default}
%\usetheme{Frankfurt}
%\usetheme{Goettingen}
%\usetheme{Hannover}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Luebeck}
%\usetheme{Madrid}
%\usetheme{Malmoe}
%\usetheme{Marburg}
%\usetheme{Montpellier}
%\usetheme{PaloAlto}
%\usetheme{Pittsburgh}
%\usetheme{Rochester}
%\usetheme{Singapore}
%\usetheme{Szeged}
%\usetheme{Warsaw}

@@

\begin{frame}{Scoring card for a theory of expectation formation}
	To look if the parameter and goodness of fit is robust to 
	\begin{enumerate}
		\item use of different moments in estimation
		\item alternative assumption about the underlying process 
		\item two-step estimation or joint-estimation 
		\item relatively fit with professionals and households  
	\end{enumerate}
\end{frame}

@@

\begin{comment}
%%%%%%%%%%%%%%%%%%%
\subsubsection{Hybird of NI and SE}

@@

\end{comment}

@@

\begin{frame}{Taking stock}
	\begin{table}[]
		\caption{A scoring card of different theory} 
		\begin{tabular}{llll}
				\hline 
			Criteria                                    & SE     & NI  & DE     \\
				\hline 
			Sensitive to moments used for estimation?   & No     & No  & No     \\
			Sensitive to the assumed inflation process? & No     & Yes & No     \\
			Sensitive to a two-step or joint estimate?  & No     & Yes & No     \\
			Sensitive to the type of agents?            & Yes    & Yes & Yes    \\
			Matching with FE                            & Yes    & Yes & Yes    \\
			Matching with disagreement                  & Yes    & No  & Unsure \\
			Matching with uncertainty                   & Unsure & No  & No    \\
			\hline 
		\end{tabular}
	\end{table}

\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{frame}{Conclusion}
	\begin{itemize}
		\item Sticky expectation (SE) augmented with stochastic volatility of inflation process matches data of inflation and expectations better than other theories for both professionals and households.
		\item Within each model,  households are more inconsistent  compared to professionals
		\item Incorporating higher moments, i.e. uncertainty,  helps ``discipline'' theories on expectation formation
		\item  Higher moments from surveys also contain useful information about the inflation dynamics itself
	\end{itemize}	
\end{frame}
